A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-05-28 manuscript/manuscript.tex
Manuscript updated

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{luacolor}
  \usepackage[soul]{lua-ul}
\else
  \usepackage{soul}

\fi
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading,figureheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Process for Emulating Comparative Oncology Trials with
Real-world Evidence Studies (ENCORE): Process Development and
Methodological Considerations for Oncology Real-World Data}
\author{}
\date{}

\begin{document}
\maketitle

@@

\ul{\textbf{Manuscript word count:}} 4,131 words / 8,000 words

\ul{\textbf{Abstract word count:}} 248 words / 250 words

@@

\ul{\textbf{Supplementary material:}} Supplementary figures and material

@@

\ul{\textbf{Analytic code sharing statement:}} Code to reproduce this
manuscript, including all figures and tables, can be found at
\url{https://github.com/janickweberpals/encore-process-manuscript}.

@@

\emph{Manuscript last updated: 2025-03-30 19:11:35.894604}

@@

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: parallel assignment
\item
  Industry-sponsored
\item
  Trial start in 2011 or later
\item
  Primary purpose was study treatment efficacy
\item
  Overall survival must be one of the endpoints reported (either as
  hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or such that challenged existing treatment
  policies)
\end{itemize}

@@

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The fundamental points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{19}
including assessment of whether critical eligibility criteria (e.g.,
biomarker status) including prognostic factors (e.g., ECOG performance
score) are measurable and whether preliminary study size is reasonable.
Lastly, trial candidates were ranked and shortlisted into primary and
runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts} and the corresponding selection process is
illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
Naturally, the majority of trials cover advanced (locally advanced,
inoperable, recurrent/progressive disease) or metastatic cancer
populations because a large proportion of drug development efforts have
recently focused on these settings. A key objective that we aim to
explore with the selected trials is to achieve a better understanding
how different disease settings (early, late), line settings
({[}neo{]}adjuvant, first line, advanced lines of therapy), therapy
protocols (monotherapy, combination therapy) and clinical
characteristics (all-comer versus complex genetic or immunological
signatures) can be emulated using RWD. If more thorough feasibility
assessments suggest that the threat of bias from mis-measurement of key
study parameters or residual confounding remains high, or that the study
size is not sufficient, then runner-up candidate trials will be
considered instead.

@@

@@

Additionally, we will provide conceptual and operationalized definitions
on how exposures, outcomes and covariates are defined in each respective
database. There will be special emphasis on how exposure, in context of
disease and line of therapy settings, and the OS outcome are emulated.
For all considered databases, OS is typically a composite measure that,
depending on the underlying database, can be derived from different
mortality sources (i.e., EHR documentation, social security death index,
obituary and other linkages). Given that not all relevant sources that
provide mortality data are synchronized and updated uniformly,
sensitivity analyses with more conservative (i.e., earlier) censoring
dates will be considered for each trial emulation to mitigate the
potential impact of ghost-time bias.\textsuperscript{26}

@@

One important aspect when emulating oncology trials is the choice and
estimation of the appropriate estimand of interest.\textsuperscript{27}
Particularly when emulating pivotal trials of paradigm-changing
treatments, multiple aspects need to be considered such as the
contemporaneity of the control cohort, the adoption rate of the novel
medical product in routine care, the magnitude of the clinical treatment
benefit and the rate in which (particularly patients in the control arm)
discontinue or cross-over to the interventional treatment, which could
lastly bias emulated treatment effects towards the null. To that end,
comprehensive data explorations will be performed as part of the
protocol development to contextualize these parameters and (if reported)
draw comparisons to the emulated trial. Examples for such standard
diagnostics are visualized in Figure~\ref{fig-initiators}. All
exploratory analyses will be conducted blinded towards the outcome to
avoid influencing study design and analytic choices.

@@

Causal analyses of observational data are often not designed with
respect to formal hypothesis testing and statistical power in the same
manner as RCTs since the number of `recruited' patients is limited by
available data and cannot be influenced.\textsuperscript{28} For this
project, however, the emulations in EHR data must have at least equal
power to the relevant trial for interpretation of the pre-specified
agreements metrics between RCT and RWD results. Since the main outcome
of interest is defined as time to all-cause mortality (OS), the
estimation of the statistical power is driven by the number of events
rather than the number of patients. To assess if the overall number of
events, unstratified by exposure, is sufficient such that a significant
difference can be detected based on the original RCT-reported hazard
ratio (HR), the statistical power (1-\(\beta\)) will be estimated using
Schoenfeld's sample-size formula for the proportional-hazards regression
model.\textsuperscript{29}

@@

To formally compare treatment effects between RCTs and their respective
emulations, we will adapt the approach of the RCT-DUPLICATE
project.\textsuperscript{9, 30} That is, for the primary endpoint of
interest (HR for OS and corresponding 95\% confidence intervals), we
will derive three qualitative agreement metrics: statistical agreement,
estimate agreement and agreement based on the standardized mean
difference (SMD). Examples are illustrated Table~\ref{tbl-metrics}.

@@

The study design for each trial emulation will be visualized as part of
the protocol using a graphical depiction of the exact measurement
windows of eligibility criteria, washout periods and covariates relative
to the cohort entry time.\textsuperscript{31}

@@

To establish an analytic cohort, missingness will be assessed in a first
iteration across patients who meet all eligibility criteria, including
those with missing values. These missing data investigations will
empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{32} We will adopt a
principled process on missing data that empirically evaluates different
aspects across partially observed covariates based on three group
diagnostics.\textsuperscript{33, 34} These diagnostics cover (1)
comparisons of patients characteristics with and without an observed
level of the partially observed covariate, (2) ability to predict
missingness given observed data, and (3) assessments if outcomes between
patients with a missing value are systematically different. Expert
domain knowledge and assumptions about the underlying missing data
structure through canonical causal diagrams\textsuperscript{35} will
additionally inform decisions regarding the inclusion or exclusion of
patients with missing values in key eligibility criteria and potential
sensitivity analyses to assess the robustness of these
decisions.\textsuperscript{36}

While the MAR assumption is a strong assumption to hold across all
considered covariates, it has been shown that especially in the context
of partially observed covariate data (as opposed to missing exposure and
outcome data), only mechanisms in which a covariate causes its own
missingness lead to critical bias (MNAR).\textsuperscript{35} Hence,
methodologies which retain patients and give the potential to adjust for
a broader set of prognostic factors (e.g., multiple
imputation\textsuperscript{37} or doubly robust
methods\textsuperscript{38}) may be preferred over complete case
analyses.

@@

Due to its ubiquity in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal hazard ratio (HR)
coefficient for the treatment comparison for OS.\textsuperscript{39} We
will also consider alternative endpoints on an absolute risk scale such
as median survival times, survival probabilities at pre-defined time
points during follow-up\textsuperscript{40} and restricted mean survival
times as secondary endpoints of interest.

For the estimation of marginal treatment effects, we will employ
propensity score methods to adjust for measured confounding between
treatment arms. The selection of important prognostic covariates will be
based on expert clinical knowledge and published literature on
prognostic scores in oncology.\textsuperscript{41} The implementation of
propensity scores in combination with multiple imputation will follow
the `\emph{within}' methodology as described by Leyrat et
al.\textsuperscript{42, 43} That is, propensity score matching or
weighting will be applied to each imputed dataset. The marginal
treatment effect will then be estimated in each imputed and matched or
weighted dataset separately and pooled into a final estimate following
Rubin's rule.\textsuperscript{44, 45} This approach has been shown to
lead to unbiased estimates across different simulated scenarios with a
sufficient estimation of the variance.\textsuperscript{42}

To asses the balance of pre-exposure covariates after matching or
weighting on each imputed dataset, the average SMD and corresponding
minimum and maximum SMD range will be visualized (see example in
Figure~\ref{fig-balance}). Covariate balance is often reasonably
considered at a SMD \textless{} 0.1.\textsuperscript{46} Further, we
will compute the average post-matching or post-weighting
C-statistics.\textsuperscript{47} In addition, we will use a published
prognostic score for OS\textsuperscript{41} as a balance measure to
visually assess if the prognostic score is balanced between treatment
arms after propensity score matching or weighting as this approach was
described to show the highest correlations with bias compared with other
balance measures and not affected by model
misspecification.\textsuperscript{48}

Similarly, survival probabilities for individual time points will be
estimated in each imputed and propensity score matched or weighted
dataset according to the Kaplan-Meier method.\textsuperscript{40} Since
survival probabilities typically do not follow normal distributions
which are required to apply Rubin's rule, these will be transformed
through a complementary log-log transformation \(log(-log(1-pr(surv)))\)
with \(pr(surv)\) denoting the survival probability at a given time
during follow-up.\textsuperscript{49, 50} The transformed survival
probabilities are then pooled across imputed datasets and individual
time points following Rubin's rule and back-transformed via
\(1-exp(-exp(qbar))\) with \(qbar\) denoting the pooled survival
probability. The median survival time can be finally determined by
extracting the time point during follow-up at which the survival
probability drops below 0.5 for the first time.

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\ifLuaTeX
  \usepackage{luacolor}
  \usepackage[soul]{lua-ul}
\else
  \usepackage{soul}
\fi

% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\KOMAoption{captions}{tableheading,figureheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Process for Emulating Comparative Oncology Trials with
Real-world Evidence Studies (ENCORE): Process Development and
Methodological Considerations for Oncology Real-World Data}
\author{}
\date{}
\begin{document}
\maketitle

@@

\ul{\textbf{Manuscript word count:}} 4,509 words / 8,000 words

\ul{\textbf{Abstract word count:}} 224 words / 250 words

@@

\ul{\textbf{Supplementary material:}} Supplementary figures, tables and
material

@@

\ul{\textbf{Analytic code sharing statement:}} Simulated examples and
code to implement analytic workflows described in this manuscript are
illustrated at
\url{https://janickweberpals.github.io/imputation-ps-workflows/} and can
be reproduced via the \texttt{encore.analytics} R package
(\url{https://github.com/janickweberpals/encore.analytics/}).

@@

\emph{Manuscript last updated: 2025-05-28 16:51:13.630823}

@@

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: parallel assignment
\item
  Industry-sponsored
\item
  Trial start in 2011 or later
\item
  Primary purpose was to study treatment efficacy
\item
  Overall survival must be one of the endpoints reported (either as
  hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or such that challenged existing treatment
  policies)
\end{itemize}

@@

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The fundamental points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{19}
involving an assessment of whether critical eligibility criteria (e.g.,
biomarker status) including prognostic factors (e.g., ECOG performance
score) are measurable, and whether preliminary study size based on a
rough estimation of the number of patients observed in the data with the
combination of treatment regimen and line of treatment is reasonable for
shortlisting the trial. Lastly, trial candidates were ranked and
shortlisted into primary and runner-up candidates based on their
clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts}. The corresponding selection process is illustrated
in the CONSORT diagrams (\textbf{Supplementary Figures 1-4}) and runner
up candidate trials are listed in \textbf{Supplementary Table 1}.
Naturally, the majority of shortliested trials cover advanced (locally
advanced, inoperable, recurrent/progressive disease) or metastatic
cancer populations because a large proportion of drug development
efforts have recently focused on these settings. A key objective that we
aim to explore with the selected trials is to achieve a better
understanding how different disease settings (early, late), line
settings ({[}neo{]}adjuvant, first line, advanced lines of therapy),
therapy protocols (monotherapy, combination therapy) and clinical
characteristics (all-comer versus complex genetic or immunological
signatures) can be emulated using RWD. If more thorough feasibility
assessments after applying all relevant eligibility criteria suggest
that the threat of bias from mis-measurement of key study parameters or
residual confounding remains high, or that the study size is not
sufficient, then runner-up candidate trials will be considered instead.

@@

\textbf{Emulating the RCT estimand}

One important aspect when emulating oncology trials is the choice and
estimation of the appropriate estimand of interest.\textsuperscript{26}
In prior RCT-DUPLICATE efforts, a ``while on treatment'' strategy was
chosen for database studies that were designed to emulate trials where
the primary trial analysis was intention-to-treat (e.g., a ``treatment
policy'' estimand). This estimand was chosen to mimic the high adherence
typically observed in large cardiovascular clinical trials in the
context of low adherence in clinical practice. However, a ``while on
treatment'' estimand would be inappropriate when emulating oncology
trials, not only because of highly informative censoring, but also
because OS outcome analyses in oncology are typically ``treatment
policy'' estimands that allow crossover.\textsuperscript{27} We plan to
focus on ``treatment policy'' estimands for the primary analyses in our
trial emulations. However, allowing crossover can dilute treatment
effects, moving point estimates toward the null. Therefore, differences
in cross-over and discontinuation rates between the trial and emulation
may still contribute to estimand differences. Recognizing these
practical challenges, we will carefully characterize the patterns of
discontinuation or cross-over in both the trials and the database
studies designed to emulate them.

@@

Additionally, we will provide conceptual and operationalized definitions
on how exposures, outcomes and covariates are defined in each respective
database. There will be special emphasis on how exposure, in context of
disease and line of therapy settings, and the OS outcome are emulated.
For all considered databases, OS is typically a composite measure that,
depending on the underlying database, can be derived from different
mortality sources (i.e., EHR documentation, social security death index,
obituary and other linkages). Given that not all relevant sources that
provide mortality data are synchronized and updated uniformly,
sensitivity analyses with more conservative (i.e., earlier) censoring
dates will be considered for each trial emulation to mitigate the
potential impact of ghost-time bias.\textsuperscript{28}

@@

Particularly when emulating pivotal trials of paradigm-changing
treatments, multiple aspects need to be considered, such as the
contemporaneity of the control cohort, the adoption rate of the novel
medical product in routine care, the magnitude of the clinical treatment
benefit and the rate in which patients discontinue or cross-over, which
could dilute effect estimates.

To that end, comprehensive data explorations will be performed as part
of the protocol development to contextualize these parameters and (if
reported) draw comparisons to the emulated trial. Examples for such
standard diagnostics are visualized in Figure~\ref{fig-initiators}. All
exploratory analyses will be conducted blinded towards the outcome to
avoid influencing study design and analytic choices.

@@

Causal analyses of observational data are often not designed with
respect to formal hypothesis testing and statistical power in the same
manner as RCTs since the number of `recruited' patients is limited by
available data and cannot be influenced.\textsuperscript{29} For this
project, however, the emulations in EHR data must have at least equal
power to the relevant trial for interpretation of the pre-specified
agreements metrics between RCT and RWD results. Since the main outcome
of interest is defined as time to all-cause mortality (OS), the
estimation of the statistical power is driven by the number of events
rather than the number of patients. To assess if the overall number of
events, unstratified by exposure, is sufficient such that a significant
difference can be detected based on the original RCT-reported hazard
ratio (HR), the statistical power (1-\(\beta\)) will be estimated using
Schoenfeld's sample-size formula for the proportional-hazards regression
model.\textsuperscript{30}

@@

To formally compare treatment effects between RCTs and their respective
emulations, we will adapt the approach of the RCT-DUPLICATE
project.\textsuperscript{9, 31} That is, for the primary endpoint of
interest (HR for OS and corresponding 95\% confidence intervals), we
will derive three qualitative agreement metrics: statistical agreement,
estimate agreement and agreement based on the standardized mean
difference (SMD). Examples are illustrated Table~\ref{tbl-metrics}.

@@

The study design for each trial emulation will be visualized as part of
the protocol using a graphical depiction of the exact measurement
windows of eligibility criteria, washout periods and covariates relative
to the cohort entry time.\textsuperscript{32}

@@

To establish an analytic cohort, missingness will be assessed in a first
iteration across patients who meet all eligibility criteria, including
those with missing values. These missing data investigations will
empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{33} We will adopt a
principled process on missing data that empirically evaluates different
aspects across partially observed covariates based on three group
diagnostics.\textsuperscript{34, 35} These diagnostics cover (1)
comparisons of patients characteristics with and without an observed
level of the partially observed covariate, (2) ability to predict
missingness given observed data, and (3) assessments if outcomes between
patients with a missing value are systematically different. Expert
domain knowledge and assumptions about the underlying missing data
structure through canonical causal diagrams\textsuperscript{36} will
additionally inform decisions regarding the inclusion or exclusion of
patients with missing values in key eligibility criteria and potential
sensitivity analyses to assess the robustness of these
decisions.\textsuperscript{37}

While the MAR assumption is a strong assumption to hold across all
considered covariates, it has been shown that especially in the context
of partially observed covariate data (as opposed to missing exposure and
outcome data), only mechanisms in which a covariate causes its own
missingness lead to critical bias (MNAR).\textsuperscript{36} Hence,
methodologies which retain patients and give the potential to adjust for
a broader set of prognostic factors (e.g., multiple
imputation\textsuperscript{38} or doubly robust
methods\textsuperscript{39}) may be preferred over complete case
analyses.

@@

Due to its ubiquity in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal hazard ratio (HR)
coefficient for the treatment comparison for OS.\textsuperscript{40} We
will also consider alternative endpoints on an absolute risk scale such
as median survival times, survival probabilities at pre-defined time
points during follow-up\textsuperscript{41} and restricted mean survival
times as secondary endpoints of interest.

For the estimation of marginal treatment effects, we will employ
propensity score methods to adjust for measured confounding between
treatment arms. The selection of important prognostic covariates will be
based on expert clinical knowledge and published literature on
prognostic scores in oncology.\textsuperscript{42} The implementation of
propensity scores in combination with multiple imputation will follow
the `\emph{within}' methodology as described by Leyrat et
al.\textsuperscript{43, 44} That is, propensity score matching or
weighting will be applied to each imputed dataset. The marginal
treatment effect will then be estimated in each imputed and matched or
weighted dataset separately and pooled into a final estimate following
Rubin's rule.\textsuperscript{45, 46} This approach has been shown to
lead to unbiased estimates across different simulated scenarios with a
sufficient estimation of the variance.\textsuperscript{43}

To asses the balance of pre-exposure covariates after matching or
weighting on each imputed dataset, the average SMD and corresponding
minimum and maximum SMD range will be visualized (see example in
Figure~\ref{fig-balance}). Covariate balance is often reasonably
considered at a SMD \textless{} 0.1.\textsuperscript{47} Further, we
will compute the average post-matching or post-weighting
C-statistics.\textsuperscript{48} In addition, we will use a published
prognostic score for OS\textsuperscript{42} as a balance measure to
visually assess if the prognostic score is balanced between treatment
arms after propensity score matching or weighting as this approach was
described to show the highest correlations with bias compared with other
balance measures and not affected by model
misspecification.\textsuperscript{49}

Similarly, survival probabilities for individual time points will be
estimated in each imputed and propensity score matched or weighted
dataset according to the Kaplan-Meier method.\textsuperscript{41} Since
survival probabilities typically do not follow normal distributions
which are required to apply Rubin's rule, these will be transformed
through a complementary log-log transformation \(log(-log(1-pr(surv)))\)
with \(pr(surv)\) denoting the survival probability at a given time
during follow-up.\textsuperscript{50, 51} The transformed survival
probabilities are then pooled across imputed datasets and individual
time points following Rubin's rule and back-transformed via
\(1-exp(-exp(qbar))\) with \(qbar\) denoting the pooled survival
probability. The median survival time can be finally determined by
extracting the time point during follow-up at which the survival
probability drops below 0.5 for the first time.
